---
title: "Process VEP results, and upload to SciDB"
output: html_document
---

PARADIGM4 INC.

This file is distributed along with the Paradigm4 Enterprise SciDB 
distribution kit and may only be used with a valid Paradigm4 contract 
and in accord with the terms and conditions specified by that contract.

Copyright (C) 2023 Paradigm4 Inc.
All Rights Reserved.

# Setup

```{r}
library(biobank)
library(tidyverse)
```

```{r}
bb <- connect(
  host = "scidb.mazetx.com",
  username = "scidbadmin",
  password = readLines("~/.scidb_pass"),
  port = 8083,
  protocol = "https"
)
```

# Prepare annotation frames and load the first part of the data

```{r}
numeric_cols <- c("chrom", "pos",
                  "cDNA_pos", "cDNA_end",
                  "CDS_pos", "CDS_end",
                  "aa_pos", "aa_end",
                  "am_pathogenicity")

load_output_file <- function(output_file) {
  suppressWarnings(
    readr::read_tsv(output_file,
                    col_types = 'cccccccccccccccccccccccccccc',
                    skip = 51) %>% 
      dplyr::rename(Uploaded_variation = `#Uploaded_variation`) %>%
      dplyr::mutate(across(everything(), ~ na_if(.x, '-'))) %>%
      dplyr::mutate(STRAND = ifelse(STRAND=='-1','-',ifelse(STRAND=='1','+',STRAND))) %>%
      tidyr::separate_wider_delim(Uploaded_variation, delim='_', names = c("chrom", "pos", "ref_alt"),
                                  too_few = "align_start") %>%
      tidyr::separate_wider_delim(ref_alt, delim='/', names = c("ref", "alt"),
                                  too_few = "align_start") %>%
      tidyr::separate_wider_delim(Amino_acids, delim = '/', names = c("aa_ref", "aa_alt"),
                                  too_few = "align_start") %>%
      tidyr::separate_wider_delim(cDNA_position, delim = '-', names = c("cDNA_pos", "cDNA_end"),
                                  too_few = "align_start") %>%
      tidyr::separate_wider_delim(CDS_position, delim = '-', names = c("CDS_pos", "CDS_end"),
                                  too_few = "align_start") %>%
      tidyr::separate_wider_delim(Protein_position, delim = '-', names = c("aa_pos", "aa_end"),
                                  too_few = "align_start") %>%
      dplyr::mutate(across(all_of(numeric_cols), as.numeric)) %>%
      dplyr::rename(geneid = Gene, list_label = Feature) %>%
      dplyr::filter(if_all(starts_with("am_"), ~ !is.na(.))) %>%
      dplyr::select(list_label, chrom, pos, ref, alt, geneid, aa_pos, aa_ref, aa_alt, starts_with("am_"))
  )
}
```


```{r}
out_dir <- "~/staging/vep_run/output"

annotation_frame_name <- "alphamissense_2023_OCT"
note <- "AlphaMissense annotations from October 2023 for all transcripts. List label is Ensembl transcript ID."

fill_annotation_frame <- function(namespace) {
  output_files <- fs::dir_ls(fs::path(out_dir, namespace), glob = "*.tsv")
  output <- load_output_file(output_files[[1]]) %>%
    as.data.frame()
  
  field_types <- setNames(sapply(output, typeof), colnames(output))
  field_types[field_types=="character"] <- "string"
  field_types[field_types %in% c("double", "integer")] <- "double"
  if(!all(field_types %in% c("string", "double"))) stop("wrong type")
  
  fields <- field_types %>%
    tibble::enframe(name = "name", value = "type") %>%
    dplyr::filter(!name %in% c("list_label","chrom","pos","ref","alt")) %>%
    as.data.frame()

  if(annotation_frame_name %in% bb$list_annotation_frames(namespace)$annotation_frame_name)
    bb$delete_annotation_frame(namespace = namespace,
                               annotation_frame = annotation_frame_name)
  
  bb$create_annotation_frame(namespace = namespace,
                             annotation_frame_name = annotation_frame_name,
                             fields = fields,
                             note = note)

  bb$load_annotation_frame(namespace = namespace,
                           annotation_frame = annotation_frame_name,
                           data = as.data.frame(output))
  
  lapply(output_files[2:length(output_files)], function(output_file) {
    print(output_file)
    output <- load_output_file(output_file) %>%
      as.data.frame()
    bb$load_annotation_frame(namespace = namespace,
                             annotation_frame = annotation_frame_name,
                             data = as.data.frame(output))
    rm(output)
  })
}

continue_annotation_frame <- function(namespace, start_index) {
  output_files <- fs::dir_ls(fs::path(out_dir, namespace), glob = "*.tsv")

  lapply(output_files[start_index:length(output_files)], function(output_file) {
    print(output_file)
    output <- load_output_file(output_file) %>%
      as.data.frame()
    bb$load_annotation_frame(namespace = namespace,
                             annotation_frame = annotation_frame_name,
                             data = as.data.frame(output))
    rm(output)
  })
}

fill_annotation_frame("UKB_200K_WES")
fill_annotation_frame("UK_BIOBANK_470K_WES_LOW_MAF")

```


# Check that uploads were successful

Reconnect as user
```{r}
bb <- connect(
  host = "scidb.mazetx.com",
  username = "mbrauer",
  password = rstudioapi::askForPassword(),
  port = 8083,
  protocol = "https"
)
```

```{r}
bb$get_annotation_frame_fields("UKB_200K_WES", "alphamissense_2023_OCT")
bb$get_annotation_frame_fields("UK_BIOBANK_470K_WES_LOW_MAF", "alphamissense_2023_OCT")
```

```{bash}
head input/UK_BIOBANK_470K_WES_LOW_MAF_vep_input_part_00
```

```{r}
bb$get_annotation_frame("UK_BIOBANK_470K_WES_LOW_MAF", "alphamissense_2023_OCT",
                        chrom = 1,
                        start = 1000000,
                        end = 2000000)
```

```{r}
bb$get_annotation_frame(WES, "vep_annotations_all", list_label = "ENST00000688131")
```

```{r}
bb$get_annotation_frame(WES, "vep_annotations_mane_select",
                        chrom = 1,
                        pos = 1054382)
```

```{r}
bb$get_annotation_frame(WES, "vep_annotations_mane_select",
                        LoF = "HC")
```
