---
title: "Paradigm4 General Practitioner (GP) Data query examples"
output: html_document
---

PARADIGM4 INC.

This file is distributed along with the Paradigm4 Enterprise SciDB 
distribution kit and may only be used with a valid Paradigm4 contract 
and in accord with the terms and conditions specified by that contract.

Copyright (C) 2010 - 2021 Paradigm4 Inc.
All Rights Reserved.

For background on the data, see http://biobank.ndph.ox.ac.uk/showcase/refer.cgi?id=591
UKBB also provided a set of mapping spreadsheets that we used to map from read codes to their meanings, ICD10 and OPCS4: http://biobank.ndph.ox.ac.uk/showcase/refer.cgi?id=592

# 0. Connecting
```{r}
library(biobank)
```

```{r}
bb <- connect(
  host = "127.0.0.1",
  username = 'scidbadmin',                  #replace with your username
  password = rstudioapi::askForPassword(), 
  port = 8083,
  protocol = "https")
bb$namespace='UK_BIOBANK'
```

Note that no GP data is currently ingested.

# 1. GP Scripts Data

This data contains records for each drug prescription, searchable by eid, drug names and codes.

## 1.1 Keyword Search

Search 'drug_name' by keyword. This returns case-insensitive keyword matches for the supplied term.
```{r}
bb$get_gp_scripts(drug_name_contains = 'lipitor')
```

The argument `drug_name_contains` supports regular expressions. You can search multiple keywords like this:
```{r}
x = bb$get_gp_scripts(drug_name_contains = '(lipitor|atorvastatin).*40')
x
```

Note the variation in drug names
```{r}
table(x$drug_name)
```

## 1.2 Other Options

Exact match for drug name. Can use a vector of multiple names:
```{r}
bb$get_gp_scripts(drug_name=c('LIPITOR tabs 40mg', 'LIPITOR 40mg tablets'))
```

Exact match for BNF code:
```{r}
bb$get_gp_scripts(bnf_code=c('0212000B0BBAAAA', '0212000B0BBABAB'))
```

Exact match for DMD code:
```{r}
bb$get_gp_scripts(dmd_code = '320030001')
```

Aggregations:
```{r}
all_lipitor_terms = bb$get_gp_scripts(namespace = 'UK_BIOBANK', 
                                  drug_name_contains = '(lipitor|atorvastatin)', 
                                  aggregate="count(*)", group_by = c("drug_name"))
all_lipitor_terms
```

Subset by data_provider:
```{r}
bb$get_gp_scripts(namespace = 'UK_BIOBANK', 
                  drug_name=all_lipitor_terms$drug_name,
                  data_provider = c(2,3))
```

Subset by eid:
```{r}
bb$get_gp_scripts(namespace = 'UK_BIOBANK', eid = c(5064325, 2374891, 2374891))
```

# 2. GP Clinical Data

The clinical data covers doctors' observations, procedures and some diagnostic measurements. The overall usage is similar to scripts but there are more terms and codes to search by.

Search by eid:
```{r}
bb$get_gp_clinical( eid=4693115 )
```

Search by ICD code / data provider. Note you can use exact match like "icd10 = 'I251'" but some read codes map to multiple ICD codes and those would not be returned
```{r}
bb$get_gp_clinical(icd10_contains = 'I25', data_provider = c(3,4))
```

Search the LKP terms for a keyword, with aggregation:
```{r}
bb$get_gp_clinical(lkp_contains = 'asthma', aggregate = "count(*)", group_by="lkp")
```

Look at procedure codes mapped to OPCS4:
```{r}
procedures = bb$get_gp_clinical(aggregate="count(*)", group_by=c("opcs4", "lkp"))
head(procedures[order(procedures$count, decreasing=T), ])
```

Look at disease codes mapped to ICD10:
```{r}
icd10_codes = bb$get_gp_clinical(aggregate="count(*)", group_by=c("icd10", "lkp"))
head(icd10_codes[order(icd10_codes$count, decreasing=T), ])
```

Get all LKP terms:
```{r}
all_lkp = bb$get_gp_clinical(aggregate = "count(*)", group_by="lkp")
all_lkp[order(all_lkp$count, decreasing = T), ]
```

Search for lkp term by exact match:
```{r}
bb$get_gp_clinical(namespace='UK_BIOBANK', lkp = 'Asymptomatic coronary heart disease')
```

Search for read3 codes:
```{r}
bb$get_gp_clinical(namespace='UK_BIOBANK', read_3=c('XaG1Q', 'G34z0'))
```

# X: Admin Section: Updating Data

Admin users may update the GP data like so. This will only work for users who have administrator privileges. First, download the updated files (gp_clinical.txt and gp_scripts.txt) from UKBB and place them on the server into a directory like ~/gp.
Make sure to use the full directory path here, like '/home/USERNAME/gp'. Do not use '~'.

```{r}
  bb$load_gp_data(namespace = 'UK_BIOBANK',
                  data_directory = '/home/my_user_name/gp')
```