---
title: "Paradigm4 Training Notebook: Covid-19 Test Data"
output: html_document
---

PARADIGM4 INC.

This file is distributed along with the Paradigm4 Enterprise SciDB 
distribution kit and may only be used with a valid Paradigm4 contract 
and in accord with the terms and conditions specified by that contract.

Copyright (C) 2010 - 2021 Paradigm4 Inc.
All Rights Reserved.

# 0. Intro

This notebook covers some example queries using COVID-19 test data.  Note that no data is currently loaded.

The UKBB data format, and therefore the API interface described in this notebook, may change in the future.

For any questions, please write to us at maze_support@paradigm4.com

# 1. Make a SciDB connection
```{r}
library(biobank)
```

```{r}
bb <- connect(
  host = "127.0.0.1",
  username = 'scidbadmin',
  password = rstudioapi::askForPassword(),
  port = 8083,
  protocol = "https")
bb$namespace='UK_BIOBANK'
```

# 2. Query COVID test data

```{r}
covid_tests = bb$get_covid_data(namespace = 'UK_BIOBANK')
head(covid_tests)
```

Each entry in this result is a "test event". Note there are multiple tests per individual
```{r}
nrow(covid_tests)
```

As of the first batch, there were, on average, about two tests conducted for each person. This is the number of unique individuals:
```{r}
length(unique(covid_tests$eid))
```

We can extract the list individuals who tested positive at least once:
```{r}
covid19_positive_eid = unique(subset(covid_tests, result==1)$eid)
length(covid19_positive_eid)
head(covid19_positive_eid)
```


# 3. Extract Phenotypic Data for these Individuals

We can pull phenotypic data for these individuals. Let's find the "age at recruitment" field
```{r}
bb$get_phenotype_fields('UK_BIOBANK', 'UKBB_45561', name_kw='age at recruitment')
```

And extract
```{r}
covid_age_at_recruitment = bb$get_phenotype_data('UK_BIOBANK', 'UKBB_45561', 
                                                 field_id=21022, 
                                                 eid=covid19_positive_eid,
                                                 value_cast='double')
head(covid_age_at_recruitment)
```

```{r}
hist(covid_age_at_recruitment$value, breaks = 30, col='blue')
```

# 4. Query Diagnoses

The field 41270 has all the ICD10 diagnoses - primary and secondary. We can extract all diagnoses for these people:
```{r}
covid19_icd10 = bb$get_phenotype_data('UK_BIOBANK', 'UKBB_45561', field_id=41270, 
                                      eid=covid19_positive_eid)
head(covid19_icd10)
```

Look at the 20 most common diagnoses:
```{r}
top_20_icd10= head(sort(table(covid19_icd10$value), decreasing = T), n=20)
top_20_icd10
```

Convert codes to meanings and look at their counts in the total population
```{r}
icd10_meanings = bb$get_icd10(namespace = 'UK_BIOBANK', code=names(top_20_icd10))
icd10_meanings$total_count = rowSums(icd10_meanings[,3:ncol(icd10_meanings)], na.rm=T)
icd10_meanings= icd10_meanings[c('coding', 'meaning', 'total_count')]
icd10_meanings= icd10_meanings[order(icd10_meanings$total_count, decreasing = T), ]
icd10_meanings
```

# 5. Query Clinical Records

Get the clinical records for these individuals
```{r}
clinical_records = bb$get_gp_clinical(namespace = 'UK_BIOBANK', eid = covid19_positive_eid)
head(clinical_records)
```

Query Medication data.
Here we use `aggregate` to create a unique entry for each eid/drug pair
```{r}
scripts = bb$get_gp_scripts(namespace = 'UK_BIOBANK', eid = covid19_positive_eid,
                            aggregate="count(*)", group_by=c("eid", "drug_name"))
head(scripts)
```

```{r}
common_drugs = data.frame( table(scripts$drug_name))
colnames(common_drugs) = c('drug_name', 'num_eids')
head(common_drugs[order(common_drugs$num_eids, decreasing = T), ], n=10)
```

# 6. Other Examples

Note you can use the `eid` argument in `bb$get_genotypes`. You can also convert testing positive for COVID-19 to a phenotype and run an association job - though the sample size may be too small for a meaningful result. We expect that the test data will be updated regularly with more test results, at which point such analyses should become more meaningful.

Reach out to maze_support@paradigm4.com with any questions!


# X. Admin Section

## 1. The test data
These commands are for admin use only.

To load COVID-19 data, as an admin user on the server. Put the covid19.txt file in a directory like '~/covid' and load like so:
```{r}
bb$load_covid_data(namespace = 'UK_BIOBANK', directory = '~/covid')
```


## 2. The MISC blood group data

This should only need to be done once. To load the COVID-19 MISC table (blood group information), read in the covid19_misc.txt file:
```{r}
misc_data = data.table::fread('~/covid/covid19_misc.txt')
misc_data$eid = as.numeric(misc_data$eid)
misc_data$blood_group = as.character(misc_data$blood_group)
misc_data
```

And load it as a phenotype set:
```{r}
bb$upload_phenotype_set(
  namespace = 'UK_BIOBANK',
  phenotype_set_name = 'COVID19_MISC', 
  data = data.frame(misc_data), 
  value_type = 'Categorical (single)', 
  description = 'Blood group from the COVID19_MISC file',
  note = 'Covid-19 Misc data (blood group information), uploaded by... YOUR_USER_NAME 1/1/14 ...'
)
```

### Check the set
```{r}
bb$list_phenotype_sets('UK_BIOBANK')
```

```{r}
bb$get_phenotype_fields('UK_BIOBANK', 'COVID19_MISC')
```

### Get the blood group data
```{r}
bb$get_phenotype_data('UK_BIOBANK', 'COVID19_MISC', title='blood_group')
```

### Delete the set, if needed
```{r}
# bb$delete_phenotype_set(namespace = 'UK_BIOBANK', phenotype_set = 'COVID19_MISC')
```
