---
title: "Paradigm4 Training Notebook: Loading Phenotype Baskets"
output: html_document
---

PARADIGM4 INC.

This file is distributed along with the Paradigm4 Enterprise SciDB 
distribution kit and may only be used with a valid Paradigm4 contract 
and in accord with the terms and conditions specified by that contract.

Copyright (C) 2010 - 2021 Paradigm4 Inc.
All Rights Reserved.

# 0. Connecting 
```{r}
library(biobank)
```
```{r}
bb <- connect(
  host = "127.0.0.1",
  username = 'scidbadmin',                  #replace with your username
  password = rstudioapi::askForPassword(), 
  port = 8083,
  protocol = "https")
bb$namespace='UK_BIOBANK'
```

# 1. Loading a phenotype basket

In this document, we'll walk through loading a phenotype basket - using the 46920_basket as an example.  You'll need write permissions to the namespace where you want to load the phenotype basket.

# 1.1.  Unpack and convert phenotype basket download

You'll need the ukbunpack and ukbconv utilities (https://biobank.ctsu.ox.ac.uk/showcase/download.cgi).  Replace paths below as appopriate.

First you'll need to upack and decrypt the .enc file.  This will take a while for a "full" phenotype basket, but is fast for the small basket we're using here.
```{r}
system("cd /staging/46920_basket_2012493/; ~/ukbunpack ukb46920.enc k50314r46920.key")
```

Then you'll need to convert the file to text format.
```{r}
system("cd /staging/46920_basket_2012493/; ~/ukbconv ukb46920.enc_ukb txt")
```

# 1.2.  Load phenotype fields & encodings

Load latest field dictionary and field encodings - important if the phenotype basket contains new fields, as 46920 did.

```{r}
bb$load_field_dictionary(namespace = 'UK_BIOBANK',
                         overwrite_if_exists = T)
```

```{r}
bb$load_encodings(namespace = 'UK_BIOBANK', overwrite_if_exists = T)
```

# 1.3.  Load phenotype basket into biobank

```{r}
bb$load_phenotype_basket(namespace = 'UK_BIOBANK',
                         phenotype_set_name = 'UKBB_46920',  
                         basket_file = '/staging/46920_basket_2012493/ukb46920.txt',
                         update_icd_counts = F)
```

```{r}
bb$list_phenotype_sets(namespace = "UK_BIOBANK")
```