---
title: "Installing Paradigm4 packages"
output: html_notebook
---

Copyright Paradigm4 2023 &copy;

Follow these instructions to install the Reveal:Biobank&trade; client package on your
local machine, to allow remote connections to SciDB. For now, use of the maze
VPN is required to connect.

## Install bioconductor dependencies

```{r echo=TRUE, results='hide'}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("IRanges")
```

Test IRanges installation:

```{r echo=TRUE, results='hide'}
library(IRanges)
```


## Install arrow

This is needed for fast data download from SciDB. It depends on (and
automatically installs) the arrow system library; that step can require
additional configuration. If you aren't able to successfully install arrow,
you can move on to the next step; most functionality in biobank will work
without arrow (though some legacy datasets require it).

```{r echo=TRUE, results='hide'}
install.packages("arrow")
```

Test arrow installation:

```{r}
library(arrow, warn.conflicts = F)
dat <- arrow_table(x = 1:3, y = c("a", "b", "c"))
dat
as.data.frame(dat)
```


## Install P4 packages and dependencies

You will need to be connected to the maze VPN for this installation, and for
subsequently accessing the database.

If prompted "do you want to install from source" or "do you want to install
dependencies", answer "y"

```{r echo=TRUE, results='hide'}
install.packages(c("scidb", "arrayop", "BurstMode", "biobank"),
  repos = c("https://cran.rstudio.com/", "http://scidb.mazetx.com/p4_r_repo/"))
```

Test biobank installation:

```{r echo=TRUE, results='hide'}
library(biobank, warn.conflicts = F)
bb <- connect(
  host = "scidb.mazetx.com",
  username = "your_username", # <- YOUR USERNAME HERE
  password = rstudioapi::askForPassword()
  # password = readline() # Use this instead if not running from RStudio
)
```

```{r}
bb$list_namespaces()
```

If you don't see the "UK_BIOBANK" namespace, contact maze_support@paradigm4.com
to be added to the security group.

```{r}
bb$get_variants("UK_BIOBANK", chrom = 1, start = 1, end = 100000, fields = c("rsid", "maf"))
```

