---
title: "Phecode Construction"
output: html_document
---

PARADIGM4 INC.

This file is distributed along with the Paradigm4 Enterprise SciDB
distribution kit and may only be used with a valid Paradigm4 contract
and in accord with the terms and conditions specified by that contract.
Copyright (C) 2023 Paradigm4 Inc.

All Rights Reserved.

# 0. About

This notebook illustrates how to assemble a PheCode workflow into a stored 
Phenotype Set. A quick PheWas example at the end shows how to use the uploaded
phenotypes.

We've run this notebook once and created the set PHECODE_UKBB_45561
For example usage, skip to the last section.

# 1. Connect

```{r}
library(biobank)
library(data.table)
```

```{r}
bb <- connect(
  host = "127.0.0.1",
  username = 'scidbadmin',
  password = rstudioapi::askForPassword(),
  port = 8083,
  protocol = "https"
)
```

# 2. Setup workspace

```{bash}
mkdir -p ~/phecodes
cd ~/phecodes
wget https://raw.githubusercontent.com/atgu/ukbb_pan_ancestry/master/data/UKB_Phecode_v1.2b1_ICD_Mapping.txt
```

```{r}
phecode_mapping = fread('~/phecodes/UKB_Phecode_v1.2b1_ICD_Mapping.txt', colClasses = 'character')
phecode_mapping
```

```{r}
bb$list_phenotype_sets("UK_BIOBANK")
```

Set the namespace/basket we are constructing PheCodes from and where we will upload them.
Note you may not have permissions to upload to `UK_BIOBANK` but you can upload to `public`

```{r}
SOURCE_NAMESPACE = 'UK_BIOBANK'
SOURCE_PHENOTYPE_SET = 'UKBB_45561'

TARGET_NAMESPACE = 'UK_BIOBANK'
TARGET_PHENOTYPE_SET_NAME = 'PHECODE_UKBB_45561'
```

Set which fields we will use. 

```{r}
# Include primary and secondary causes of death, cancer codes, external causes,
# and main and secondary diagnoses. Here, rather than including primary and 
# secondary diagnostic fields individually, we leverage fields 41270 and 41271.
# These fields aggregate unique ICD codes for each participant from both the 
# primary and secondary fields.
ICD10_FIELDS = c(
  40001, # death primary
  40002, # death secondary 
  40006, # cancer
  41201, # external
  41270) # diagnoses (primary + secondary)
ICD9_FIELDS = c(
  40013, # cancer
  41271) # diagnoses (primary + secondary)
```

```{r}
all_codes <- bb$get_phenotype_data(
  namespace = SOURCE_NAMESPACE,
  phenotype_set = SOURCE_PHENOTYPE_SET,
  field_id = c(ICD10_FIELDS, ICD9_FIELDS),
  minimal = T # we don't care about linking to dates, so don't need instance/index
)
all_codes
```


```{r}
setDT(all_codes)
setnames(all_codes, "value", "code")
all_codes[, field_id := NULL]
setkey(all_codes, "code")
all_codes
```


Download the sex data from the namespace
```{r}
sex_data_raw = bb$get_phenotype_data(
  namespace = SOURCE_NAMESPACE,
  phenotype_set = SOURCE_PHENOTYPE_SET,
  field_id = c(
    31,     # Sex
    22001), # Genetic sex
  all_eids = T
)
setDT(sex_data_raw)
encoding <- bb$get_encoding(SOURCE_NAMESPACE, encoding_id = 9)
setDT(encoding)
sex_data_decoded = encoding[sex_data_raw, c("eid", "field_id", "meaning"), on = "value"]
sex_data_decoded
```

```{r}
unique(phecode_mapping$sex)
```

```{r}
sex_data <- dcast(sex_data_decoded, eid ~ field_id, value.var = "meaning")
sex_data
```

```{r}
setnames(sex_data, old = c("31", "22001"), new = c("sex", "genetic_sex"))
sex_data
```


How many individuals are missing a genetic sex determination?
```{r}
table(sex_data$genetic_sex, useNA = 'always')
```

What are the counts for sex?
```{r}
table(sex_data$sex, useNA = "always")
```


How many mismatches are there between sex and genetic sex (for individuals
for which we have both)?

```{r}
sex_data[sex != genetic_sex]
```

```{r}
sex_data[sex == genetic_sex, agreed_sex := sex]
```

```{r}
sex_data[is.na(sex) | is.na(genetic_sex) | sex != genetic_sex]
```

```{r}
sex_data
```

```{r}
table(sex_data$agreed_sex, useNA = "always")
```

# 3. Assemble the PheCodes as a data frame

We use the first column for eid (numeric) and add one numeric (1,0,NA) column 
for each code. When using this method, local RAM usage can be a concern. At 
~1500 phecodes, the frame uses about 7GB, which is not much for our primary 
server.

```{r}
phecodes_frame <- bb$get_participants(
    namespace= SOURCE_NAMESPACE, 
    phenotype_set = SOURCE_PHENOTYPE_SET
)
setDT(phecodes_frame, key = "eid")
phecodes_frame
```

```{r}
#add cases for each PheCode
for( i in 1:nrow(phecode_mapping) ) {
  phecode_id <- phecode_mapping$phecode[i]
  #initialize all to 0
  phecodes_frame[, (phecode_id) := 0L]
  icd_codes <- strsplit(phecode_mapping$icd_codes[i], ",")[[1]]
  case_data <- all_codes[code %in% icd_codes, eid]
  phecodes_frame[eid %in% case_data, (phecode_id) := 1L]
  if (i %% 20 == 0) {
    print(sprintf("%04d / %04d", i, nrow(phecode_mapping)))
  }
}
```

```{r}
head(phecodes_frame[, 1:5])
```

```{r}
ones <- sum(phecodes_frame[, 2:length(phecodes_frame)], na.rm = T)
zeros <- sum(1 - phecodes_frame[, 2:length(phecodes_frame)], na.rm = T)
nas <- sum(is.na(phecodes_frame[, 2:length(phecodes_frame)]))
print(ones)
print(zeros)
print(nas)
print(ones + zeros + nas)
print((length(phecodes_frame) - 1) * nrow(phecodes_frame))
```

exclude sibling PheCode controls (set to NA)

```{r}
for (i in 1:nrow(phecode_mapping)) {
  phecode_id <- phecode_mapping$phecode[i]
  codes_to_exclude <- strsplit(phecode_mapping$exclude_phecodes[i], ",")[[1]]
  # Build exclusion mask of rows with 1 for any of the codes to exclude
  mask <- rowSums(phecodes_frame[, ..codes_to_exclude], na.rm = T) > 0 &
    phecodes_frame[[phecode_id]] == 0L
  phecodes_frame[mask, (phecode_id) := NA]
  
  if (i %% 20 == 0) {
    print(sprintf("%04d / %04d", i, nrow(phecode_mapping)))
  }
}
```

```{r}
head(phecodes_frame[, 1:10])
```

```{r}
ones <- sum(phecodes_frame[, 2:length(phecodes_frame)], na.rm = T)
zeros <- sum(1 - phecodes_frame[, 2:length(phecodes_frame)], na.rm = T)
nas <- sum(is.na(phecodes_frame[, 2:length(phecodes_frame)]))
print(ones)
print(zeros)
print(nas)
```

exclude males or females from sex-specific PheCodes
```{r}
not_female <- sex_data[is.na(agreed_sex) | agreed_sex == "Male", eid]
not_male <- sex_data[is.na(agreed_sex) | agreed_sex == "Female", eid]
```

```{r}
length(not_female)
length(not_male)
```


```{r}
for (i in 1:nrow(phecode_mapping)) {
  if (i %% 20 == 0) {
    print(sprintf("%04d / %04d", i, nrow(phecode_mapping)))
  }
  
  phecode_id <- phecode_mapping$phecode[i]

  sex_condition = phecode_mapping$sex[i]
  if (sex_condition == 'both_sexes') {
    next
  } else if (sex_condition == 'females') {
    # Exclude Male and Undetermined participants
    phecodes_frame[eid %in% not_female, (phecode_id) := NA]
  } else if (sex_condition == 'males') {
    # Exclude Female and Undetermined participants
    phecodes_frame[eid %in% not_male, (phecode_id) := NA]
  } else {
    stop(paste("Unexpected sex_condition:", sex_condition))
  }
}
```

```{r}
ones <- sum(phecodes_frame[, 2:length(phecodes_frame)], na.rm = T)
zeros <- sum(1 - phecodes_frame[, 2:length(phecodes_frame)], na.rm = T)
nas <- sum(is.na(phecodes_frame[, 2:length(phecodes_frame)]))
print(ones)
print(zeros)
print(nas)
```

```{r}
head(phecodes_frame[, 1:10])
```

# 4. Upload to the database

```{r}
titles = colnames(phecodes_frame)[2:ncol(phecodes_frame)]
head(titles)
```

```{r}
all(titles == phecode_mapping$phecode)
```

```{r}
titles = gsub('_$', '',
         gsub('_+', '_',
         gsub('[^[:alnum:]_]', '_',
         paste("phc", phecode_mapping$phecode, phecode_mapping$description, sep = "_")
)))
head(titles)
```

```{r}
max(nchar(titles))
titles[which.max(nchar(titles))]
```

```{r}
setnames(phecodes_frame, c("eid", titles))
```

```{r}
head(phecodes_frame[, 1:10])
```

```{r}
descriptions = paste0(phecode_mapping$phecode, ": ", phecode_mapping$description)
notes = paste0('Constructed from ICD 9/10 codes ', phecode_mapping$icd_codes, 
               ' excluding phecodes ',phecode_mapping$exclude_phecodes)
```

```{r}
head(descriptions)
head(notes)
```

```{r}
setDF(phecodes_frame)
```

```{r}
overall_note <- paste0('Phecodes generated from ',
                SOURCE_NAMESPACE, '.', SOURCE_PHENOTYPE_SET, ' fields ',
                paste0(c(ICD10_FIELDS, ICD9_FIELDS), collapse=','))
overall_note
```

```{r}
ones <- sum(phecodes_frame[2:length(phecodes_frame)], na.rm = T)
zeros <- sum(1 - phecodes_frame[2:length(phecodes_frame)], na.rm = T)
nas <- sum(is.na(phecodes_frame[2:length(phecodes_frame)]))
print(ones)
print(zeros)
print(nas)
print(ones + zeros + nas)
print((length(phecodes_frame) - 1) * nrow(phecodes_frame))
```

```{r}
print(length(phecodes_frame) - 1)
print(nrow(phecode_mapping))
print(nrow(phecodes_frame))
print(nrow(bb$get_participants(SOURCE_NAMESPACE, SOURCE_PHENOTYPE_SET)))
```

```{r}
bb$list_phenotype_sets("UK_BIOBANK")
```

Upload. Takes about 20 min to upload and arrange.
```{r}
bb$upload_phenotype_set(
  namespace = TARGET_NAMESPACE,
  phenotype_set_name = TARGET_PHENOTYPE_SET_NAME, 
  note = overall_note, 
  data = phecodes_frame, 
  value_type = rep('Categorical (single)', length(descriptions)),
  description = descriptions,
  notes = notes, 
  remove_na = FALSE,
  use_build = TRUE,
  build_value = 0
)
```


# 5 Test some example queries

```{r}
bb$namespace <- TARGET_NAMESPACE
bb$list_phenotype_sets()
```

```{r}
bb$get_phenotype_fields(phenotype_set = TARGET_PHENOTYPE_SET_NAME)
```

```{r}
bb$get_phenotype_fields(phenotype_set = TARGET_PHENOTYPE_SET_NAME, name_kw = "renal_failure")
```

```{r}
bb$get_phenotype_data(phenotype_set = TARGET_PHENOTYPE_SET_NAME, field_id = 1102, minimal = T)
```

