---
title: "Phecode Construction"
output: html_document
---

PARADIGM4 INC.

This file is distributed along with the Paradigm4 Enterprise SciDB
distribution kit and may only be used with a valid Paradigm4 contract
and in accord with the terms and conditions specified by that contract.
Copyright (C) 2023 Paradigm4 Inc.

All Rights Reserved.

# 0. About

This notebook illustrates how to assemble a PheCode workflow into a stored 
Phenotype Set. A quick PheWas example at the end shows how to use the uploaded
phenotypes.

We've run this notebook once and created the set PHECODE_AGE_UKBB_45561
For example usage, skip to the last section.

# 1. Connect

```{r}
library(biobank)
library(data.table)
```

```{r}
bb <- connect(
  host = "127.0.0.1",
  username = 'scidbadmin',
  password = rstudioapi::askForPassword(),
  port = 8083,
  protocol = "https"
)
```

# 2. Setup workspace

```{bash}
mkdir -p ~/phecodes
cd ~/phecodes
wget https://raw.githubusercontent.com/atgu/ukbb_pan_ancestry/master/data/UKB_Phecode_v1.2b1_ICD_Mapping.txt
```

```{r}
phecode_mapping = fread('~/phecodes/UKB_Phecode_v1.2b1_ICD_Mapping.txt', colClasses = 'character')
phecode_mapping
```

```{r}
bb$list_phenotype_sets("UK_BIOBANK")
```

Set the namespace/basket we are constructing PheCodes from and where we will upload them.
Note you may not have permissions to upload to `UK_BIOBANK` but you can upload to `public`

```{r}
SOURCE_NAMESPACE = 'UK_BIOBANK'
SOURCE_PHENOTYPE_SET = 'UKBB_45561'

TARGET_NAMESPACE = 'UK_BIOBANK'
TARGET_PHENOTYPE_SET_NAME = 'PHECODE_WITH_AGES_UKBB_45561'
```

Set which fields we will use. 

```{r}
# Include primary and secondary causes of death, cancer codes, external causes,
# and main and secondary diagnoses. Here, rather than including primary and 
# secondary diagnostic fields individually, we leverage fields 41270 and 41271.
# These fields aggregate unique ICD codes for each participant from both the 
# primary and secondary fields.
# Field 41201 does not have age data linked, so we will not use it.
ICD10_FIELDS = c(
  40001, # death primary
  40002, # death secondary 
  40006, # cancer
  41270) # diagnoses (primary + secondary)
ICD9_FIELDS = c(
  40013, # cancer
  41271) # diagnoses (primary + secondary)
```

# 3. Collate ICD and age data

```{r}
all_codes <- bb$get_phenotype_data(
  namespace = SOURCE_NAMESPACE,
  phenotype_set = SOURCE_PHENOTYPE_SET,
  field_id = c(ICD10_FIELDS, ICD9_FIELDS)
  # Keep instance and index information for linking to dates
)
all_codes
```

```{r}
ages <- bb$get_phenotype_data(
  namespace = SOURCE_NAMESPACE,
  phenotype_set = SOURCE_PHENOTYPE_SET,
  field_id = c(40007, 40008)
)
ages
```

```{r}
setDT(ages)
ages[, age := as.numeric(value)]
ages[, value := NULL]
ages[, title := NULL]
```

```{r}
ages
```

```{r}
year_of_birth <- bb$get_phenotype_data(
  namespace = SOURCE_NAMESPACE,
  phenotype_set = SOURCE_PHENOTYPE_SET,
  field_id = c(34),
  minimal = T
)
year_of_birth
```

```{r}
month_of_birth <- bb$get_phenotype_data(
  namespace = SOURCE_NAMESPACE,
  phenotype_set = SOURCE_PHENOTYPE_SET,
  field_id = c(52),
  minimal = T
)
month_of_birth
```

```{r}
setDT(year_of_birth)
setDT(month_of_birth)
year_of_birth[, year := as.numeric(value)]
year_of_birth[, field_id := NULL]
year_of_birth[, value := NULL]
month_of_birth[, month := as.numeric(value)]
month_of_birth[, field_id := NULL]
month_of_birth[, value := NULL]
year_of_birth
month_of_birth
```

```{r}
dob <- merge(year_of_birth, month_of_birth, by = "eid")
dob
```

```{r}
dob[, numeric_dob := year + (month / 12)]
dob
```

```{r}
anyNA(dob)
```

```{r}
dob[, year := NULL]
dob[, month := NULL]
dob
```

```{r}
date_fields <- bb$get_phenotype_data(
  namespace = SOURCE_NAMESPACE,
  phenotype_set = SOURCE_PHENOTYPE_SET,
  field_id = c(41280, 41281)
)
date_fields
```

```{r}
setDT(date_fields)
date_fields[, title := NULL]
date_fields[, date := as.Date(value)]
date_fields[, year := as.numeric(format(date, "%Y"))]
date_fields[, month := as.numeric(format(date, "%m"))]
date_fields[, numeric_date := year + (month / 12)]
date_fields
```

```{r}
date_fields[, c("value", "date", "year", "month") := NULL]
date_fields
```

```{r}
date_fields_age_converted <- merge(date_fields, dob, by = "eid", all.x = T, all.y = F)
date_fields_age_converted
```

```{r}
anyNA(date_fields_age_converted)
```

```{r}
date_fields_age_converted[, age := numeric_date - numeric_dob]
date_fields_age_converted
```

Sanity check
```{r}
min(date_fields_age_converted$age)
max(date_fields_age_converted$age)
```

```{r}
date_fields_age_converted[, c("numeric_date", "numeric_dob") := NULL]
```

```{r}
ages
date_fields_age_converted
```

```{r}
all_ages <- rbind(ages, date_fields_age_converted)
all_ages
```

```{r}
setDT(all_codes)
all_codes[, title := NULL]
all_codes
```

```{r}
anyNA(all_ages)
anyNA(all_codes)
```

```{r}
all_codes[field_id == 40001 | field_id == 40002, age_field_id := 40007] # Death
all_codes[field_id == 40006 | field_id == 40013, age_field_id := 40008] # Cancer
all_codes[field_id == 41270, age_field_id := 41280] # Diagnoses (ICD10)
all_codes[field_id == 41271, age_field_id := 41281] # Diagnoses (ICD9)
all_codes
```

```{r}
setnames(all_codes, "value", "code")
all_codes
```

```{r}
all_codes[field_id == 40002, index := 0] # Secondary causes of death all happened on the same date as primary
```

```{r}
setnames(all_ages, "field_id", "age_field_id")
all_ages
```

```{r}
codes_with_ages <- merge(all_codes, all_ages, by = c("eid", "age_field_id", "instance", "index"))
codes_with_ages
```

```{r}
anyNA(codes_with_ages)
```


Don't care about which specific field data came from any more

```{r}
codes_with_ages[, c("field_id", "age_field_id", "instance", "index") := NULL]
setkey(codes_with_ages, code) # Sort now to make filtering by code fast
codes_with_ages
```

# 4. Collate sex data

Download the sex data from the namespace
```{r}
sex_data_raw = bb$get_phenotype_data(
  namespace = SOURCE_NAMESPACE,
  phenotype_set = SOURCE_PHENOTYPE_SET,
  field_id = c(
    31,     # Sex
    22001), # Genetic sex
  all_eids = T
)
setDT(sex_data_raw)
encoding <- bb$get_encoding(SOURCE_NAMESPACE, encoding_id = 9)
setDT(encoding)
sex_data_decoded = encoding[sex_data_raw, c("eid", "field_id", "meaning"), on = "value"]
sex_data_decoded
```

```{r}
unique(phecode_mapping$sex)
```

```{r}
sex_data <- dcast(sex_data_decoded, eid ~ field_id, value.var = "meaning")
sex_data
```

```{r}
setnames(sex_data, old = c("31", "22001"), new = c("sex", "genetic_sex"))
sex_data
```


How many individuals are missing a genetic sex determination?
```{r}
table(sex_data$genetic_sex, useNA = 'always')
```

What are the counts for sex?
```{r}
table(sex_data$sex, useNA = "always")
```


How many mismatches are there between sex and genetic sex (for individuals
for which we have both)?

```{r}
sex_data[sex != genetic_sex]
```

```{r}
sex_data[sex == genetic_sex, agreed_sex := sex]
```

```{r}
sex_data[is.na(sex) | is.na(genetic_sex) | sex != genetic_sex]
```

```{r}
sex_data
```

```{r}
table(sex_data$agreed_sex, useNA = "always")
```

# 5. Collate censor age (age at death, or age at end of data collection)

```{r}
censor_age <- bb$get_phenotype_data(
  namespace = SOURCE_NAMESPACE,
  phenotype_set = SOURCE_PHENOTYPE_SET,
  field_id = 40007, # Age at death
  instance = 0, index = 0,
  all_eids = T
)
censor_age
```

```{r}
setDT(censor_age, key = "eid")
censor_age[, age := as.numeric(value)]
censor_age[, c("title", "field_id", "instance", "index", "value") := NULL]
censor_age
```

```{r}
all_dates <- bb$get_phenotype_data(
  namespace = SOURCE_NAMESPACE,
  phenotype_set = SOURCE_PHENOTYPE_SET,
  field_id = c(
    40000, # date of death
    40005, # cancer
    41280, # icd10
    41281 # icd9
  ),
  minimal = T
)
```

```{r}
all_dates
```

```{r}
latest_date <- max(as.Date(all_dates$value))
latest_date
```

```{r}
latest_date_num <- as.numeric(format(latest_date, "%Y")) +
  (as.numeric(format(latest_date, "%m")) / 12)
latest_date_num
```

```{r}
dob
```

```{r}
censor_age[!(eid %in% dob$eid)]
```
```{r}
6025289 %in% codes_with_ages$eid
```
```{r}
bb$get_phenotype_data(
  namespace = SOURCE_NAMESPACE,
  phenotype_set = SOURCE_PHENOTYPE_SET,
  field_id = c(22200, 52, 53, 21003, 21022),
  eid = 6025289
)
```
We don't seem to have any age or daterecords for this single eid; we'll exclude
them from this set (since they'll end up with 0 for all phecodes but NA
for the censor age, and thus would introduce an NA censor age into all
associations).


```{r}
censor_age <- merge(censor_age, dob, by = "eid", all.x = T)
```

```{r}
censor_age
```

```{r}
censor_age[is.na(age), age := latest_date_num - numeric_dob]
censor_age
```

```{r}
censor_age[, numeric_dob := NULL]
```

```{r}
censor_age
censor_age[is.na(age)]
key(censor_age)
```

```{r}
eids_with_no_data <- censor_age[is.na(age), eid]
eids_with_no_data
```

```{r}
sex_data <- sex_data[!(eid %in% eids_with_no_data)]
censor_age <- censor_age[!(eid %in% eids_with_no_data)]
sex_data
censor_age
```


# 6. Assemble the PheCodes as a data frame

We use the first column for eid (numeric) and add one numeric (1,0,NA) column 
for each code. When using this method, local RAM usage can be a concern. At 
~1500 phecodes, the frame uses about 7GB, which is not much for our primary 
server.

```{r}
phecodes_frame <- bb$get_participants(
    namespace= SOURCE_NAMESPACE, 
    phenotype_set = SOURCE_PHENOTYPE_SET
)
setDT(phecodes_frame, key = "eid") # Order by eid for efficient inserts

# Exclude the ageless EID:
phecodes_frame <- phecodes_frame[!(eid %in% eids_with_no_data)]
key(phecodes_frame)
phecodes_frame
```


```{r}
#add cases for each PheCode
for( i in 1:nrow(phecode_mapping) ) {
  phecode_id <- phecode_mapping$phecode[i]
  # initialize all to 0
  phecodes_frame[, (phecode_id) := 0L]
  # initialize age to censor age
  phecodes_frame[, paste0(phecode_id, "_age") := censor_age$age]
  icd_codes <- strsplit(phecode_mapping$icd_codes[i], ",")[[1]]
  case_data <- codes_with_ages[code %in% icd_codes, .(age = min(age)), keyby = eid]
  # Since we sorted both these by eid, this can be an in-place insert rather than a join
  phecodes_frame[eid %in% case_data$eid, (phecode_id) := 1L]
  phecodes_frame[eid %in% case_data$eid, paste0(phecode_id, "_age") := case_data$age]
  if (i %% 20 == 0) {
    print(sprintf("%04d / %04d", i, nrow(phecode_mapping)))
  }
}
```

The "no non-missing arguments to min; returning Inf" warning is spurious; it
occurs when there are no cases of a phecode (and thus case_data is empty), but
no `Inf` values are actually inserted.

```{r}
head(phecodes_frame[, 1:5])
```

```{r}
anyNA(phecodes_frame)
```

```{r}
num_phecodes <- nrow(phecode_mapping)
num_phecodes == (length(phecodes_frame) - 1) / 2
num_phecodes
```

```{r}
phecode_indices <- (1:num_phecodes) * 2
age_indices <- phecode_indices + 1
head..phecode_indices
tail(age_indices)
length(phecodes_frame)
head(colnames(phecodes_frame)[phecode_indices])
head(colnames(phecodes_frame)[age_indices])
tail(colnames(phecodes_frame)[phecode_indices])
tail(colnames(phecodes_frame)[age_indices])
```

```{r}
ones <- sum(phecodes_frame[, ..phecode_indices], na.rm = T)
zeros <- sum(1 - phecodes_frame[, ..phecode_indices], na.rm = T)
nas <- sum(is.na(phecodes_frame[, ..phecode_indices]))
print(ones)
print(zeros)
print(nas)
print(ones + zeros + nas)
print(num_phecodes * nrow(phecodes_frame))
```

exclude sibling PheCode controls (set to NA)

```{r}
for (i in 1:nrow(phecode_mapping)) {
  phecode_id <- phecode_mapping$phecode[i]
  codes_to_exclude <- strsplit(phecode_mapping$exclude_phecodes[i], ",")[[1]]
  # Build exclusion mask of rows with 1 for any of the codes to exclude
  mask <- rowSums(phecodes_frame[, ..codes_to_exclude], na.rm = T) > 0 &
    phecodes_frame[[phecode_id]] == 0L
  phecodes_frame[mask, (phecode_id) := NA]
  phecodes_frame[mask, paste0(phecode_id, "_age") := NA]
  
  if (i %% 20 == 0) {
    print(sprintf("%04d / %04d", i, nrow(phecode_mapping)))
  }
}
```

```{r}
head(phecodes_frame[, 1:10])
```

```{r}
ones <- sum(phecodes_frame[, ..phecode_indices], na.rm = T)
zeros <- sum(1 - phecodes_frame[, ..phecode_indices], na.rm = T)
nas <- sum(is.na(phecodes_frame[, ..phecode_indices]))
print(ones)
print(zeros)
print(nas)
```

exclude males or females from sex-specific PheCodes
```{r}
not_female <- sex_data[is.na(agreed_sex) | agreed_sex == "Male", eid]
not_male <- sex_data[is.na(agreed_sex) | agreed_sex == "Female", eid]
```

```{r}
length(not_female)
length(not_male)
```


```{r}
for (i in 1:nrow(phecode_mapping)) {
  if (i %% 20 == 0) {
    print(sprintf("%04d / %04d", i, nrow(phecode_mapping)))
  }
  
  phecode_id <- phecode_mapping$phecode[i]

  sex_condition = phecode_mapping$sex[i]
  if (sex_condition == 'both_sexes') {
    next
  } else if (sex_condition == 'females') {
    # Exclude Male and Undetermined participants
    phecodes_frame[eid %in% not_female, c(phecode_id, paste0(phecode_id, "_age")) := NA]
  } else if (sex_condition == 'males') {
    # Exclude Female and Undetermined participants
    phecodes_frame[eid %in% not_male, c(phecode_id, paste0(phecode_id, "_age")) := NA]
  } else {
    stop(paste("Unexpected sex_condition:", sex_condition))
  }
}
```

```{r}
ones <- sum(phecodes_frame[, ..phecode_indices], na.rm = T)
zeros <- sum(1 - phecodes_frame[, ..phecode_indices], na.rm = T)
nas <- sum(is.na(phecodes_frame[, ..phecode_indices]))
print(ones)
print(zeros)
print(nas)
```

```{r}
head(phecodes_frame[, 1:10])
```

# 7. Upload to the database

```{r}
head(colnames(phecodes_frame))
tail(colnames(phecodes_frame))
```

```{r}
code_titles <- phecode_mapping$phecode
all(phecode_mapping$phecode == colnames(phecodes_frame)[phecode_indices])
head(phecode_mapping$phecode)
```


```{r}
code_titles <-
  gsub('_$', '',
  gsub('_+', '_',
  gsub('[^[:alnum:]_]', '_',
  paste("phc", phecode_mapping$phecode, phecode_mapping$description, sep = "_")
)))
head(code_titles)
```

```{r}
max(nchar(code_titles))
code_titles[which.max(nchar(code_titles))]
```

```{r}
age_titles <- paste0("age_", gsub('[^[:alnum:]_]', '_', phecode_mapping$phecode))
head(age_titles)
```

```{r}
all_titles <- c("eid", rbind(code_titles, age_titles))
head(all_titles)
tail(all_titles)
```

```{r}
setnames(phecodes_frame, all_titles)
```

```{r}
head(phecodes_frame[, 1:10])
```

```{r}
code_descriptions = paste0(phecode_mapping$phecode, ": ", phecode_mapping$description)
code_notes = paste0('Constructed from ICD 9/10 codes ', phecode_mapping$icd_codes, 
               ' excluding phecodes ', phecode_mapping$exclude_phecodes)
```

```{r}
age_descriptions = paste("Age at first occurance of", phecode_mapping$phecode, "or death or end of data collection")
age_notes = paste("If diagnosed with any code comprising", phecode_mapping$phecode, "age at the first such diagnosis. If a deceased control, age at death. If a non-deceased control, age at end of data collection. NA if excluded.")
```

```{r}
descriptions <- c(rbind(code_descriptions, age_descriptions))
notes <- c(rbind(code_notes, age_notes))
```


```{r}
head(descriptions)
head(notes)
```

```{r}
setDF(phecodes_frame)
```

```{r}
overall_note <- paste0('Phecodes generated from ',
                SOURCE_NAMESPACE, '.', SOURCE_PHENOTYPE_SET, ' fields ',
                paste0(c(ICD10_FIELDS, ICD9_FIELDS), collapse=', '),
                '; ages generated from fields 40007, 40008, 41280, 41281, 34, 52; ',
                'sex exclusions based on consensus of 31 and 22001')
overall_note
```

```{r}
ones <- sum(phecodes_frame[phecode_indices], na.rm = T)
zeros <- sum(1 - phecodes_frame[phecode_indices], na.rm = T)
nas <- sum(is.na(phecodes_frame[phecode_indices]))
age_nas <- sum(is.na(phecodes_frame[age_indices]))
print(ones)
print(zeros)
print(nas)
print(age_nas)
print(ones + zeros + nas)
print(((length(phecodes_frame) - 1) / 2) * nrow(phecodes_frame))
```

```{r}
print(length(phecodes_frame) - 1)
print(nrow(phecode_mapping))
print(nrow(phecodes_frame))
print(nrow(bb$get_participants(SOURCE_NAMESPACE, SOURCE_PHENOTYPE_SET)))
```

```{r}
bb$list_phenotype_sets("UK_BIOBANK")
```

```{r}
TARGET_NAMESPACE
TARGET_PHENOTYPE_SET_NAME
```

```{r}
value_types <- rep_len(c('Categorical (single)', 'Continuous'), length(descriptions))
head(value_types)
tail(value_types)
length(value_types)
```


Upload. Takes a few hours to upload and arrange.
```{r}
bb$upload_phenotype_set(
  namespace = TARGET_NAMESPACE,
  phenotype_set_name = TARGET_PHENOTYPE_SET_NAME, 
  note = overall_note, 
  data = phecodes_frame, 
  value_type = value_types,
  description = descriptions,
  notes = notes, 
  remove_na = FALSE
)
```


# 8. Test some example queries

```{r}
bb$namespace <- TARGET_NAMESPACE
bb$list_phenotype_sets()
```

```{r}
bb$get_phenotype_fields(phenotype_set = TARGET_PHENOTYPE_SET_NAME)
```

```{r}
bb$get_phenotype_fields(phenotype_set = TARGET_PHENOTYPE_SET_NAME, name_kw = "renal_failure")
```

```{r}
bb$get_phenotype_data(phenotype_set = TARGET_PHENOTYPE_SET_NAME, field_id = c(2204, 2205), minimal = T)
```

