---
title: "Paradigm4 Training Notebook: Phenotype Set Construction"
output: html_document
---

PARADIGM4 INC.

This file is distributed along with the Paradigm4 Enterprise SciDB 
distribution kit and may only be used with a valid Paradigm4 contract 
and in accord with the terms and conditions specified by that contract.

Copyright (C) 2010 - 2020 Paradigm4 Inc.
All Rights Reserved.

Update: 20201130 - new version of biobank forbids phenotype field titles with '.' in the name. Adjusted our ALL_ sets accordingly.

# 0. Connecting
```{r}
library(biobank)
```

NOTE: Replace with your username and save a copy of this notebook to your home directory
```{r}
bb <- connect(
  host = "127.0.0.1",
  username = 'scidbadmin',
  password = rstudioapi::askForPassword(),
  port = 8083,
  protocol = "https")
bb$namespace='UK_BIOBANK'
```


## 1. Unrelated White British Covariates

### 1.1. Remove Heterozygosity Outliers
Note we use `value=NULL` to retrieve entries where the field is NOT populated
```{r}
all_participants = bb$get_participants(namespace='UK_BIOBANK')
non_outliers <- bb$get_phenotype_data(namespace='UK_BIOBANK', 
                                      phenotype_set = "UKBB_43361",
                                      field_id = 22027, value = NULL)
nrow(all_participants) - nrow(non_outliers)
```

```{r}
eids = non_outliers['eid']
nrow(eids)
```

### 1.2. Remove: “Submitted.Gender” != “Inferred.Gender”
```{r}
submitted_gender <- bb$get_phenotype_data(namespace='UK_BIOBANK', phenotype_set = "UKBB_43361", field_id = 31)[c("eid", "value")]
inferred_gender <- bb$get_phenotype_data(namespace='UK_BIOBANK', phenotype_set = "UKBB_43361", field_id = 22001)[c("eid", "value")]
eids_filter <- merge(submitted_gender, inferred_gender)["eid"]
eids <- merge(eids, eids_filter)
nrow(eids)
```

### 1.3 Remove: “Putative.sex.chromosome.aneuploidy” == 1
```{r}
eids_filter <- bb$get_phenotype_data(namespace='UK_BIOBANK', phenotype_set = "UKBB_43361", field_id = 22019, value=NULL)
eids <- merge(eids, eids_filter)
nrow(eids)
```

### 1.4. Subset to British White Ancestry
```{r}
eids_filter <- bb$get_phenotype_data(namespace='UK_BIOBANK', phenotype_set = "UKBB_43361", field_id = 22006, value_prefix = 1)["eid"]
eids <- merge(eids, eids_filter)
nrow(eids)
```

### 1.5. Remove relatives
```{r}
eids_filter <- bb$get_phenotype_data(namespace='UK_BIOBANK', phenotype_set = "UKBB_43361", field_id = 22020, value_prefix = 1)["eid"]
eids <- merge(eids, eids_filter)
nrow(eids)
```

### 1.6. Remove: “excluded.from.kinship.inference” ==1 and “excess.relatives” == 1
```{r}
eids_filter <- bb$get_phenotype_data(namespace='UK_BIOBANK', phenotype_set = "UKBB_43361", field_id = 22021, value = c(0, 1))["eid"]
eids <- merge(eids, eids_filter)
nrow(eids)
```

Thus we arrive at a "White British Unrelated" subpopulation.

### 1.7 Age at enrollment
```{r}
eids_filter <- bb$get_phenotype_data(namespace='UK_BIOBANK',phenotype_set = "UKBB_43361", 
                                     field_id = 21022, value_cast='double')[c("eid", "value")]
eids_filter$age <- eids_filter$value
eids_filter$value <- NULL
eids <- merge(x = eids, y = eids_filter, all.x = TRUE)
eids = eids[, c('eid', 'age')]
eids
```

### 1.8 Sex
```{r}
eids_filter <- bb$get_phenotype_data(namespace='UK_BIOBANK', phenotype_set = "UKBB_43361", field_id = 31,
                                     value_cast='double')[c("eid", "value")]
eids_filter$sex <- eids_filter$value
eids_filter$value <- NULL
eids <- merge(x = eids, y = eids_filter, all.x = TRUE)
eids
```

### 1.9 First 10 principal components
Note we use the data.table function dcast to convert principal components to separate columns.
```{r}
eids_filter <- bb$get_phenotype_data(phenotype_set = "UKBB_43361", field_id = 22009,
                                     index_range = c(1, 10), value_cast="float")
eids_filter <- reshape2::dcast(eids_filter, eid ~ index)
field_names <- names(eids_filter)
names(eids_filter) <- c("eid", paste0("pcomp_", field_names[2:length(field_names)]))
eids <- merge(x = eids, y = eids_filter, all.x = TRUE)
eids
```

This will be a data frame we upload
```{r}
was_phenos=eids
was_phenos
```

### 1.10 Upload

Note: your account may not have permissions to modify namespace 'UK_BIOBANK' but you can upload sets to namespace 'public'
```{r}
bb$delete_phenotype_set('UK_BIOBANK', 'WHITE_BRITISH_UNRELATED_COVARIATES')
```

```{r}
bb$upload_phenotype_set(
  namespace = 'UK_BIOBANK', 
  phenotype_set_name = 'WHITE_BRITISH_UNRELATED_COVARIATES', 
  note = 'Age, Sex and first 10 Principal Components across the unrelated White British subpopulation', 
  data = was_phenos[c('eid', 'age', 'sex', 'pcomp_1', 'pcomp_2', 'pcomp_3', 'pcomp_4', 'pcomp_5', 'pcomp_6', 'pcomp_7', 'pcomp_8', 'pcomp_9', 'pcomp_10')],
  value_type = rep('Continuous', 12),
  description = c('Age at recruitment',
                  'Sex', 
                  'Principal Component 1',
                  'Principal Component 2',
                  'Principal Component 3',
                  'Principal Component 4',
                  'Principal Component 5',
                  'Principal Component 6',
                  'Principal Component 7',
                  'Principal Component 8',
                  'Principal Component 9',
                  'Principal Component 10'),
  notes = c('From UKBB_43361, field 21022',
            'From UKBB_43361, field 31',
            'From UKBB_43361, field 22009',
            'From UKBB_43361, field 22009',
            'From UKBB_43361, field 22009',
            'From UKBB_43361, field 22009',
            'From UKBB_43361, field 22009',
            'From UKBB_43361, field 22009',
            'From UKBB_43361, field 22009',
            'From UKBB_43361, field 22009',
            'From UKBB_43361, field 22009',
            'From UKBB_43361, field 22009')
)

```

```{r}
bb$list_phenotype_sets('UK_BIOBANK')
```

```{r}
bb$get_phenotype_fields('UK_BIOBANK', 'WHITE_BRITISH_UNRELATED_COVARIATES')
```

## 2. Example Continuous and Categorical Phenotypes

### 2.1 Start with all participants
```{r}
all_participants = bb$get_participants(namespace='UK_BIOBANK')
phenos = all_participants
phenos
```

### 2.2 Add E78: high cholesterol
```{r}
e78_high_cholesterol = bb$get_phenotype_data(
  namespace = 'UK_BIOBANK',
  phenotype_set = 'UKBB_43361',
  field_id = 41270, 
  value_prefix = 'E78'
)
phenos$e78_high_cholesterol = 0
phenos$e78_high_cholesterol [ phenos$eid %in% e78_high_cholesterol$eid ] = 1
phenos
```

### 2.3 Add self-reported high cholesterol
```{r}
sr_high_cholesterol = bb$get_phenotype_data(
  namespace = 'UK_BIOBANK',
  phenotype_set = 'UKBB_43361',
  field_id = 20002, 
  value_prefix = '1473'
)
phenos$sr_high_cholesterol = 0
phenos$sr_high_cholesterol [ phenos$eid %in% sr_high_cholesterol$eid ] = 1
phenos
```

### 2.4 Add a field for combined (self-reported or e78)
```{r}
phenos$high_cholesterol_combined = as.numeric(phenos$e78_high_cholesterol | phenos$sr_high_cholesterol)
phenos
```

### 2.5 Add continuous field for LDL direct, inverse-rank normal transformed
```{r}
ldl_direct = bb$get_phenotype_data(
  namespace = 'UK_BIOBANK',
  phenotype_set = 'UKBB_43361',
  field_id = 30780, 
  value_cast = 'double',
  instance = 0, #important: don't capture repeat measures
  index = 0
)
ldl_direct = ldl_direct[, c('eid', 'value')]
colnames(ldl_direct) = c('eid', 'ldl_direct')
ldl_direct$ldl_direct = RNOmni::RankNorm(ldl_direct$ldl_direct)
phenos = merge( phenos, ldl_direct, all.x = T)
phenos
```

### 2.6 Upload
```{r}
bb$delete_phenotype_set('UK_BIOBANK', 'EXAMPLE_GWAS_PHENOS')
```

```{r}
bb$upload_phenotype_set(
  namespace = 'UK_BIOBANK', 
  phenotype_set_name = 'EXAMPLE_GWAS_PHENOS', 
  note = 'A few example phenotypes related to high cholesterol, across all participants.', 
  data = phenos[c('eid', 'e78_high_cholesterol', 'sr_high_cholesterol', 'high_cholesterol_combined', 'ldl_direct')],
  value_type = c('Categorical (single)', 'Categorical (single)', 'Categorical (single)', 'Continuous'),
  description = c('Hypercholesterolemia ICD10 Code E78', 
                  'Self-reported high cholesterol',
                  'ICD10 Code E78 or Self-reported high cholesterol',
                  'LDL Direct continuous measurement, rank-normalized'),
  notes = c('From UKBB_43361, field 41270, prefix E78',
            'From UKBB_43361, field 20002, value 1473',
            'From UKBB_43361, field 41270, prefix E78 OR field 20002, value 1473',
            'From UKBB_43361, field 30780, instance 0')
)

```

```{r}
bb$list_phenotype_sets('UK_BIOBANK')
```
