---
title: "Perform Cox PH analysis of a PheCode against OmicsPred Metabolon PGS scores"
output: html_document
---

BEGIN_COPYRIGHT

PARADIGM4 INC.
This file is distributed along with the Paradigm4 Enterprise SciDB 
distribution kit and may only be used with a valid Paradigm4 contract 
and in accord with the terms and conditions specified by that contract.

Copyright (C) 2010 - 2023 Paradigm4 Inc.
All Rights Reserved.

END_COPYRIGHT

# Connect to biobank

```{r setup}
library(biobank)
library(data.table)
library(survival)
bb <- connect(
  host = "127.0.0.1",
  username = 'dgealow',
  password = rstudioapi::askForPassword(),
  port = 8083,
  protocol = "https"
)
```


```{r}
bb$list_phenotype_sets("UK_BIOBANK")
```

```{r}
bb$list_phenotype_sets("public")
```

# Download data

## Covariates

### Genotyping array

```{r}
bb$get_phenotype_fields("UK_BIOBANK", "UKBB_45561", name_kw = "batch")
```


```{r}
gt_array <- bb$get_phenotype_data(
  "UK_BIOBANK",
  "UKBB_45561",
  field_id = 22000,
  value_cast = "double",
  minimal = TRUE
)
```

```{r}
setDT(gt_array, key = "eid")
gt_array
```

```{r}
anyNA(gt_array)
```

```{r}
gt_array[, field_id := NULL]
gt_array[value > 0, array := 1L]
gt_array[value < 0, array := 0L]
```

```{r}
gt_array
```

```{r}
gt_array[, value:= NULL]
```

### sex and genetic PCs

```{r}
bb$get_phenotype_fields("UK_BIOBANK", "WHITE_BRITISH_UNRELATED_COVARIATES_45561")
```

```{r}
sex_and_PCs <- bb$get_phenotype_data(
  "UK_BIOBANK",
  "WHITE_BRITISH_UNRELATED_COVARIATES_45561",
  field_id = 1:11,
  value_cast = "double"
)
```

```{r}
setDT(sex_and_PCs)
sex_and_PCs
```

```{r}
anyNA(sex_and_PCs)
```

```{r}
sex_and_PCs <- dcast(sex_and_PCs, eid ~ title)
sex_and_PCs
```

```{r}
setnames(sex_and_PCs, old = paste0("pcomp_", 1:10), new = paste0("pc", 1:10))
```

```{r}
anyNA(sex_and_PCs)
```

### Merge together

```{r}
covariates <- merge(gt_array, sex_and_PCs, by = "eid")
setcolorder(covariates, c("eid", "sex", "array", paste0("pc", 1:10)))
covariates
```

```{r}
covariates[, sex := as.integer(sex)]
```

```{r}
anyNA(covariates)
```


## Phecode

```{r}
phecode_fields <- bb$get_phenotype_fields("UK_BIOBANK", "PHECODE_WITH_AGES_UKBB_45561")
```

```{r}
phecode_fields
```

Let's take chronic kidney disease as an example.

```{r}
bb$get_phenotype_fields("UK_BIOBANK", "PHECODE_WITH_AGES_UKBB_45561", name_kw = "CKD")
```

```{r}
phecode_field_id <- 2204
```

```{r}
phecode_data <- bb$get_phenotype_data(
  "UK_BIOBANK",
  "PHECODE_WITH_AGES_UKBB_45561",
  field_id = c(phecode_field_id, phecode_field_id + 1),
  value_cast = "double",
  minimal = TRUE
)
```

```{r}
setDT(phecode_data)
phecode_data
```

```{r}
phecode_data <- dcast(phecode_data, eid ~ field_id)
phecode_data
```

```{r}
phecode_data[is.na(`2204`) != is.na(`2205`)]
```

```{r}
phecode_data <- na.omit(phecode_data)
phecode_data
```

```{r}
setnames(phecode_data, old = as.character(c(phecode_field_id, phecode_field_id + 1)), new = c("phe", "age"))
phecode_data[, phe := as.integer(phe)]
```

```{r}
phecode_data
```

```{r}
all_pheno <- merge(phecode_data, covariates, by = "eid")
all_pheno
```


## OmicsPred

```{r}
op_fields <- bb$get_phenotype_fields("public", "OmicsPred")
```

```{r}
setDT(op_fields)
metabolon_fields <- op_fields[grepl("metab", description, ignore.case = T)]
metabolon_fields
```

Clean up to save memory.
*WARNING*: this will delete any other objects you have in your environment!
```{r}
#do.call(rm, as.list(setdiff(ls(), c("metabolon_fields", "all_pheno", "bb"))))
```

```{r}
cat("Starting...\n")
start_time <- proc.time()
score <- lapply(metabolon_fields$title, function(title){
  cat(sprintf("\rRunning %s ; %.1f min elapsed", title, (proc.time() - start_time)["elapsed"]/60))
  
  # Download and munge OmicsPred PGS score
  pgs_data <- bb$get_phenotype_data("public", "OmicsPred", title = title, value_cast = "double", minimal = T)
  setDT(pgs_data, key = "eid")
  pgs_data[, field_id := NULL]
  setnames(pgs_data, "value", "op_score")
  
  # Merge data together
  all_data <- merge(pgs_data, all_pheno, by = "eid")
  if (anyNA(all_data)) stop("NAs in all_data")
  
  # Normalize and adjust data
  all_data[, eid := NULL] # don't need EID anymore
  cols_to_scale <- c("op_score", paste0("pc", 1:10))
  setcolorder(all_data, cols_to_scale)
  ## Center and scale score and PCs
  scaled_cols <- as.data.table(scale(all_data[, ..cols_to_scale]))
  all_data[, (cols_to_scale) := scaled_cols]
  ## Adjust score for PCs
  score_lm <- lm(
    paste("op_score ~", paste0("pc", 1:10, collapse = " + ")),
    all_data
  )
  all_data[, op_score := score_lm$residuals]
  ## Normalize score again
  all_data[, op_score := scale(op_score)]
  
  # Perform Cox PH regression
  fit <- coxph(
    Surv(age, phe) ~ op_score +
           pc1 + pc2 + pc3 + pc4 + pc5 + pc6 + pc7 + pc8 + pc9 + pc10 +
           array + strata(sex),
    all_data
  )
  summ <- summary(fit)
  stopifnot(summ$coefficients["op_score", "exp(coef)"] == summ$conf.int["op_score", "exp(coef)"])
  return(data.table(
    trait = title,
    hr = summ$coefficients["op_score", "exp(coef)"],
    hr_low = summ$conf.int["op_score", "lower .95"],
    hr_high = summ$conf.int["op_score", "upper .95"],
    pvalue = summ$coefficients["op_score", "Pr(>|z|)"]
  ))
})
end_time <- proc.time()
cat("\nDone!\n")
```

```{r}
scores_dt <- rbindlist(scores)
```

```{r}
scores_dt
```

```{r}
scores_dt[order(pvalue)]
```

Write out the scores to a file
(commented to avoid accidental clobber)
```{r}
#fwrite(scores_dt, "/staging/omicspred_phewas/phc_585.3.csv")
```

```{r}
end_time - start_time
```

```{r}
(end_time - start_time)["elapsed"] / 60
```

Took about 98 minutes to run.

```{r}
qqman::qq(scores_dt$pvalue)
```


